---
title: "Replication of Study Wolff et al. (2009) and Kim & Marsden (2026) on the Initiator Hypothesis in Turkish"
author: "Aslı Kuzgun"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html: 
    toc: true
    toc_depth: 3
---

## Introduction

Wolff et al. (2009) proposes the initiator hypothesis that states in languages with morphological case (e.g., Korean, Turkish), the external argument of a clause [i.e., the subject] must be able to self energize. This restriction results in the following contrast:
The husband surprised the wife. (Highly acceptable in English/Mandarin).
The ring surprised the wife. (Highly acceptable in English/Chinese, but significantly less natural for Korean speakers).                   (Wolff et al., 2009) 
Kim & Marsden (2026) further extends this hypothesis to psych-verbs and shows that the difference persists between the case marking (Korean) and non-case marking languages (English and Mandarin). The Initiator Hypothesis predicts Turkish speakers to align with Korean speakers since they are both case marking languages. 

This study aims to test the initiator hypothesis with Turkish speakers. Since there is no study that confirms this relative contrast replicates in a case marking language other than Korean, I will employ the Turkish translations of the material in Wolff et. al. to show that the Initiator hypothesis is relevant to Turkish, and compare the data to the Korean and English speakers in Wolff et al. (2009). Lastly, the experimental design will follow  Kim & Marsden (2026) who aimed to replicate Wolff et al. (2009) with psych verbs. 

References:
Kim, J., & Marsden, H. (2026). Cross-linguistic constraints on subjecthood in causative psych verbs: An experimental investigation of Korean, Mandarin Chinese and English. Glossa Psycholinguistics, 5(1).

Wolff, P., Jeon, G. H., & Li, Y. (2009). Causers in English, Korean, and Chinese and the individuation of events. Language and Cognition, 1(2), 167-196.


## Methods

### Power Analysis

### The statistical models used in the original studies.

Wolff et al. uses F-tests for the main effects of language and energy generation of the subjects.
Kim & Marsden (2026) use Regression coefficients in their report for main effects of subjecthood.

A power analysis for the Wolff et al. study yields an f2 = 1.185, which requires 12 participants for 80% power. 
I think this is because Wolff et al. is a multilingual study and having 18 participants per language is a small sample size that might have cause an increase in the f values.


A power analysis for Kim & Marsden (2026) has an f2 of 0.41 and this yields 22 participants for 80% power.


### Planned Sample
Wolff et al. (2009) has 18 speakers for each language (i.e., English, Mandarin, and Korean) while Kim & Marsden (2026) employs 44 Korean speakers.

Considering the previous studies and the power analysis I plan a sample size of 60 native Turkish speakers. The participants will be preselected based on whether they are native speakers and if they use Turkish in their daily lives, to eliminate any attrition and intervention bias.

### Materials

This study employs the stimuli developed for Experiment 1 by Wolff et al. (2009). The sentences in Wolff et al. (2009) will be translated into Turkish.  Wolff et al. (2009) has 30 sentences with inanimate external arguments. Half of these sentences have external arguments that are mostly able to generate their own energy, as shown in (1).

The air conditioner cooled the room. (Wolff et al., 2009)

The external arguments of the other half have external arguments that are unable to generate their own energy, (2).

The knife cut the bread. (Wolff et al., 2009)

Kim & Marsden (2026) has 32 target sentences with 16 animate and 16 inanimate external arguments. This study also has 24 filler items. 

Since I replicate the design in Kim & Marsden (2026), with the materials in Wolff et al. (2009), I will have the direct translations of the 30 sentences in Wolff et al. (2009), along with 20 filler items with different acceptability conditions. There will be 50 sentences in total. The fillers will serve as attention checkers. 

### Procedure	

In this experiment, participants are shown sentences one by one on Prolific*. They rate the naturalness of sentences on a 5-point scale (0 to 4), where 4 means "completely natural" and 0 means "completely unnatural." Participants will be explicitly told to base their judgments on their linguistic intuition. The entire process should take 10–15 minutes.

* I instead used my socials to recruit participants to ensure native speaker status and demographic representation.
### Analysis Plan

Following Kim and Marsden (2006), I will implement the model using Cumulative Link Mixed Models (CLMMs) in R. The primary fixed effect will be Condition (self-energizing vs. non-self-energizing subjects). I will use Estimated Marginal Means (EMMs) to conduct pairwise comparisons between the two conditions. I will attempt to fit the maximal model, however, if it fails to converge (as in Kim and Marsden 2006), I will simplify the model by including random slopes for items only. I will use sum-coding to assess how much each condition deviates from the average rating.

### Data cleaning

Participant data will be screened based on responses to grammatical and ungrammatical fillers. Participants will be excluded if they fail to show a clear differentiation between these categories. In addition, any participant with incomplete data or suspicious response patterns (e.g., selecting the same rating for every item) will be removed from the analysis.

### Differences from Original Study

Unlike Kim and Marsden, this study focuses on a single language (Turkish), therefore, there won’t be a group condition.
After analyzing the Turkish data, I will compare the results to the Wolff et al. study.
This will have to include some changes because Wolff et al. used a 0–100 scale and F-tests, unlike the design I am following.  Therefore, the comparison with their study will be conceptual and focus on whether Turkish speakers show a significantly larger drop in acceptability for non-initiating subjects than English or Chinese speakers, aligning with Korean speakers.


Explicitly describe known differences in sample, setting, procedure, and analysis plan from original study.  The goal, of course, is to minimize those differences, but differences will inevitably occur.  Also, note whether such differences are anticipated to make a difference based on claims in the original article or subsequent published research on the conditions for obtaining the effect.

### Methods Addendum (Post Data Collection)

You can comment this section out prior to final report with data collection.

#### Actual Sample
  A total of116 participants recruited in total. The participants are all native speakers of Turkish and are based in Turkey. The data was collected through the author's socials (either by direct request or through social media). The age group varies between 20-50 years old. The sample is filtered based on the attention checkers and the participants who didn't finish the experiment. Participants who didn't finish the experiment are excluded from the analysis and those who had less than 80% accuracy on the attention checkers are also excluded.

  After the filtering process, I ended up with 101 participants.


#### Differences from pre-data collection methods plan
Kim and Marsden did not filter their data based on attention checkers, but I decided to do so to ensure the quality of the data. Also, I had a larger sample size than the original study, which might have increased the power of the study and the chances of finding a significant effect. Compared to the Wolff et al. study, I used a different scale (0-4 instead of 0-100) and a different analysis method (CLMMs instead of F-tests), which might have affected the results. Moreover, my sample size was significantly larger than that of the original study. Lastly, none of the original studies include Turkish speakers, which the current study focuses on, so the comparison with the original studies will be conceptual.
## Results


### Data # R
```{r}
# libraries
library(tidyverse)
library(readxl)
library(ordinal)
library(emmeans)
library(osfr)
library(ggdist)
library(ggplot2)
library(readr)
library(dplyr)
library(purrr)
library(stringr)
# show dplyr warnings
dplyr::last_dplyr_warnings()

```

```{r}

# final participant counts
experiment_only <- if ("task" %in% names(df_clean)) df_clean %>% filter(!is.na(task)) else df_clean

participants_tbl <- experiment_only %>%
  filter(!is.na(participant_id)) %>%
  group_by(participant_id) %>%
  summarise(n_responses = n(), n_missing_rt = sum(is.na(rt_num)), .groups = "drop") 

# Store the debugging list
cleaning_audit <- list(
  problematic_values = check, 
  problems_summary = problems_preview, 
  n_missing_after = n_missing_after,
  counts = participants_tbl
)

# Print a readable output
cat("Initial data import complete. Total participants recorded:", nrow(participants_tbl), "\n")
```


#### Prepare data for analysis - create columns etc.

```{r}


modeling_input <- experiment_only %>%
  mutate(
    rt = as.numeric(rt),
    rating = as.numeric(rating)
  )

# Verify item types (critical vs fillers)
cat("Item type distribution:\n")
print(table(modeling_input$item_type))

# Show the structure
glimpse(modeling_input)

# How many critical items each person has
check_counts <- modeling_input %>%
  filter(item_type == "critical") %>%
  group_by(participant_id) %>%
  summarise(n_items = n())

# This shows the distribution
hist(check_counts$n_items)
```
#### Data exclusion / filtering
```{r}

# Threshold for attention check accuracy: %80 success on fillers
ACCURACY_THRESHOLD <- 0.80

modeling_data <- modeling_input %>%
  # Exclude pilot participants
  filter(participant_id != "byt4qop83n") %>%
  
  # Passing criteria for all filler types
  mutate(passed_check = case_when(
    # Grammatical fillers: Expect high ratings (3 or 4)
    item_type == "filler" & condition == "grammatical" ~ if_else(rating >= 3, 1, 0),
  
    
    # Violation fillers: Expect low ratings (0, 1, or 2) for all types
    item_type == "filler" & condition %in% c("semantic_violation", "case_marking_error", "tense_mismatch") ~ if_else(rating <= 2, 1, 0),
    

    # Exclude critical items from accuracy calculation
    item_type == "critical" ~ NA_real_,
    
    # Default to 0 if none of the above
    TRUE ~ 0
  )) %>%
  
  # Accuracy and total counts per participant
  group_by(participant_id) %>%
  mutate(
    n_critical = sum(item_type == "critical"),
    n_fillers = sum(!is.na(passed_check)),
    filler_accuracy = sum(passed_check, na.rm = TRUE) / n_fillers
  ) %>%
  
  # Applying the filters
  filter(n_critical >= 20 & filler_accuracy >= ACCURACY_THRESHOLD) %>%
  ungroup() %>%
  
  # Format after filtering
  filter(item_type == "critical") %>%
  mutate(
    rating = factor(as.character(rating), levels = c("0", "1", "2", "3", "4"), ordered = TRUE),
    condition = factor(as.character(condition))
  )

# Final sample size
cat("Final sample size after exclusion:", n_distinct(modeling_data$participant_id), "participants.\n")


#CLMM (Cumulative Link Mixed Model) following Kim & Marsden (2026)
# Apply sum contrasts and run model
contrasts(modeling_data$condition) <- contr.sum(length(levels(modeling_data$condition)))
model <- clmm(rating ~ condition + (1|participant_id) + (1|item_id), data = modeling_data)
summary(model)


```

### linear model with numeric ratings
```{r}
# Convert the ordered factor back to numeric for lmer
model <- lmer(as.numeric(as.character(rating)) ~ condition + (1 | participant_id) + (1 | item_id), 
              data = modeling_data)

summary(model)
```
### Data visualization
```{r}
library(ggplot2)
library(ggdist)

# Filter data for critical items
plot_data <- modeling_data %>%
  mutate(rating_num = as.integer(as.character(rating)))

means_data <- plot_data %>%
  group_by(condition) %>%
  summarise(mean_val = mean(rating_num, na.rm = TRUE))

ggplot(plot_data, aes(x = condition, y = rating_num, fill = condition)) +
  # Dots represent individual data points (distribution)
 stat_dots(
    side = "left", 
    justification = 1.1, 
    binwidth = 0.2,          
    dotsize = 1.5,
    overflow = "compress",   # prevent dots from going beyond the plot area
    alpha = 0.7
  ) +
  # Bars represent the mean and confidence intervals (summary statistics)
  stat_pointinterval(
    point_interval = mean_qi, 
    .width = c(.66, .95),
    position = position_nudge(x = 0.1)
  ) +
    # Adding the text labels
  geom_text(
    data = means_data,
    aes(x = condition, y = mean_val, label = round(mean_val, 2)),
    vjust = -1,        # Position above the mean point
    size = 4,          
    fontface = "plain"
  ) +
  scale_y_continuous(breaks = 0:4, limits = c(-0.5, 4.5)) +
  scale_fill_manual(values = c("#56B4E9", "#E69F00")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Acceptability Ratings",
    subtitle = "Distribution and Mean Confidence Intervals",
    x = "Condition",
    y = "Rating (0-4)"
  ) +
  guides(fill = "none")
  ```
### Confirmatory analysis

The analyses as specified in the analysis plan.  

```{r}

summary(model)
```

#### Estimated Marginal Means

```{r}
#### Estimated Marginal Means
library(emmeans)

# Calculate EMMs from the model 
em_results <- emmeans(model, ~ condition)

# Print the table
knitr::kable(em_results, digits = 3, caption = "Estimated Marginal Means by Condition")
```

#### Plot: mean rating by condition
```{r}
# Create a numeric column 
rating_values <- as.numeric(as.character(modeling_data$rating))
modeling_data$rating_num <- rating_values

# Check for NAs in the new numeric column
print(paste("Number of NAs in rating_num:", sum(is.na(modeling_data$rating_num))))

# Summarise data for plotting
summary_table <- modeling_data %>%
  group_by(condition) %>%
  summarise(
    Mean = mean(rating_num, na.rm = TRUE),
    SD = sd(rating_num, na.rm = TRUE),
    N = n()
  )

knitr::kable(summary_table, digits = 2, caption = "Descriptive Statistics by Condition")
```


### Descriptive statistics (computed)


### The distribution of Acceptability ratings by condition
```{r}
library(ggplot2)
library(dplyr)
library(paletteer)

# Filtering for critical items 
plot_data <- modeling_data %>%
  filter(condition %in% c("high_energy", "low_energy")) %>%
  group_by(condition, rating_num) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(condition) %>%
  mutate(percentage = count / sum(count) * 100)

# Colorblind-friendly palette
my_colors <- paletteer_d("colorblindr::OkabeIto")[1:2]

# Plot the distribution of ratings by condition
ggplot(plot_data, aes(x = factor(rating_num), y = percentage, fill = condition)) +
  geom_col(width = 0.7, color = "black") +
  # Add labels with exact percentages
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, 
            size = 3.5) +
  facet_grid(condition ~ .) + 
  scale_x_discrete(limits = c("0", "1", "2", "3", "4")) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + 
  theme_minimal() +
  labs(
    title = "Distribution of Acceptability Ratings by Condition",
    x = "Acceptability Rating (0-4)",
    y = "Percentage of Responses (%)"
  ) +
  scale_fill_manual(values = my_colors) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 12, face = "bold")
  )
```

#### Binary acceptability percentages by condition

```{r}
library(dplyr)
library(knitr)

# Define acceptability based on ratings (only 3 or 4 are acceptable)
acceptability_data <- modeling_data %>%
  mutate(
    rating_num = as.integer(as.character(rating)),
    is_acceptable = if_else(rating_num >= 3, "Acceptable", "Unacceptable")
  )

# Check the decisions on acceptability classification
print("Spot check: First 10 rows of acceptability classification:")
print(head(acceptability_data[, c("rating_num", "is_acceptable")], 10))

# Calculate percentages
acceptability_summary <- acceptability_data %>%
  group_by(condition, is_acceptable) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(condition) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  # Show the acceptable percentages 
  filter(is_acceptable == "Acceptable") %>%
  select(condition, Acceptability_Percentage = percentage)

# 3. Print the table
kable(acceptability_summary, digits = 1, caption = "Percentage of Acceptable Ratings (3 or 4) by Condition")
```
```{r}
stats <- read.csv("results/statistics.csv", stringsAsFactors = FALSE)
if (nrow(stats) > 0) knitr::kable(stats)

```


## Discussion

### Summary of Replication Attempt

The aim of this conceptual replication was to test the initiator hypothesis in Turkish, which is a case-marking language, and compare the results to the original study by Wolff et al. (2009) and the replication by Kim & Marsden (2026). Results show that the initiator hypothesis holds for Turkish speakers. That is, Turkish speakers rated sentences with non-self-energizing subjects as significantly less acceptable than those with self-energizing subjects. That is, Turkish speakers rated sentences with non-self-energizing subjects as significantly less acceptable than those with self-energizing subjects (p < 0.001). That is, high self-energizing subjects (e.g., "the air conditioner") were rated acceptable 65.5% of the time, while low self-energizing subjects (e.g., "the knife") were rated acceptable only 41.8% of the time. This pattern aligns with the findings from Korean speakers in Wolff et al. (2009), supporting the generalizability of the initiator hypothesis to Turkish. 

### Commentary


Add open-ended commentary (if any) reflecting (a) insights from follow-up exploratory analysis, (b) assessment of the meaning of the replication (or not) - e.g., for a failure to replicate, are the differences between original and present study ones that definitely, plausibly, or are unlikely to have been moderators of the result, and (c) discussion of any objections or challenges raised by the current and original authors about the replication attempt.  None of these need to be long.

Wolff et al. (2009) reports no significant interaction between energy generation and acceptability for English and Mandarin speakers, while Korean speakers show a significant drop in acceptability for non-self-energizing subjects. This study shows that Turkish speakers also show a significant drop in acceptability for non-self-energizing subjects. This provides further support for the initiator hypothesis and suggests that the presence of morphological case marking in a language may be associated with constraints on subjecthood related to energy generation. Since neither study reports effect sizes, I cannot compare the magnitude of the effect in Turkish speakers to that of Korean speakers. However, the calculated effect size for Turkish ($d = 0.5$) confirms a medium effect size, which confirms the significant drop in the acceptability of non-self-energizing subjects among Turkish speakers. 

# Cohen's d for the difference between self-energizing and non-self-energizing subjects in Turkish speakers is calculated as follows:
```{r}
# Calculate means and SDs per condition
stats <- modeling_data %>%
  group_by(condition) %>%
  summarise(
    mean = mean(as.numeric(as.character(rating))),
    sd = sd(as.numeric(as.character(rating))),
    n = n()
  )

# Calculate Cohen's d
m1 <- stats$mean[1]
m2 <- stats$mean[2]
sd1 <- stats$sd[1]
sd2 <- stats$sd[2]
n1 <- stats$n[1]
n2 <- stats$n[2]

pooled_sd <- sqrt(((n1 - 1)*sd1^2 + (n2 - 1)*sd2^2) / (n1 + n2 - 2))
cohens_d <- abs(m1 - m2) / pooled_sd

cat("Turkish Effect Size (Cohen's d):", cohens_d)
```